Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Seminar April 27, CEDAR, Umeå}

### 2018-01-30 particle3.Rmd
first commit

```{r readalldata}
source("R/get.file.R")
skel <- get.file("skel")
##
skume <- get.file("skume")
##
foli <- get.file("foli")
skel$urban <- FALSE
vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
                    FUN = sum)
```

@@

```{r getske4060}
cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##

@@

##
vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
vb4064$age <- factor(vb4064$age)
##vb4064$hisclass <- vb4064$hisclass40
```

```{r hazcalw}
source("R/getPHglm2.R")
source("R/getPHglm.R")
##nper <- 4
nper <- length(labb)
resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
            nper, labb)
```

```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Women 40-64", ylim = c(0, 1))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Men 40-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswtab,cap="Table"}
vb4064$social_class <- vb4064$hisclass
vb4064$marital_status <- vb4064$civst
vb4064$urban_or_rural <- vb4064$urban
vb4064$age_group <- vb4064$age
fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb4064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
dr
##knitr::kable(dr)
##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
##source("R/plotED.R")
source("R/getPHglm.R")
resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
            nper, labb)

>>>>

```{r readalldata}
source("R/get.file.R")
##skel <- get.file("skel")
load("data/skel.rda")
##
##skume <- get.file("skume")
##
load("data/skume.rda")
foli <- get.file("foli")
skel$urban <- FALSE
vb0Tab <- rbind(skel, skume[, names(skel)], foli[, names(skel)])
vb0Tab$civst <- relevel(vb0Tab$civst, ref = "married")
vb0Tab <- aggregate(vb0Tab[, c("event", "exposure")],
                    vb0Tab[, c("sex", "age", "period", "cohort", "civst", "hisclass", "urban")],
                    FUN = sum)
vb0Tab <- vb0Tab[!(vb0Tab$age %in% c(30, 35)), ]
```

@@

```{r getske4060}
cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0m <- "Median age at death for a 40-year-old by social class and period, women (top) and men (bottom)."
scap0m <- "Median age at death for a 40-year old by social class."

@@

##
vb40 <- vb0Tab
vb40$age <- factor(vb40$age)
##vb40$hisclass <- vb40$hisclass40
```

```{r hazcalw}
source("R/pgetPHglm2.R")
source("R/getPHglm.R")
##nper <- 4
nper <- length(labb)
resw <- pgetPHglm2(vb40[vb40$sex == "female", ],
            nper, labb)
resm <- pgetPHglm2(vb40[vb40$sex == "male", ],
            nper, labb)
```

```{r meli, fig.cap = cap0m, fig.height = 8, out.height = "14cm"}
oldpar <- par(mfrow = c(2, 1))
plot(1:13, resw[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Women", ylab = "Age", xlab = "", xlim = c(1, 15.5))
axis(1, at = 1:13, labels = labb, las = 2)
axis(2, at = c(40, 50, 60, 70), las = 1)
abline(h = c(50, 60, 70), lty = 3)
box()
lines(1:13, resw[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
lines(1:13, resw[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
lines(1:13, resw[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
abline(h = 40)
legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
## Men:
plot(1:13, resm[1, ], type = "b", col = scol[1], axes = FALSE, ylim = c(40, 75), main = "Men", ylab = "Age", xlab = "", xlim = c(1, 15.5))
axis(1, at = 1:13, labels = labb, las = 2)
axis(2, at = c(40, 50, 60, 70), las = 1)
box()
lines(1:13, resm[2, ], type = "b", lty = 2, pch = 2, col = scol[2])
lines(1:13, resm[3, ], type = "b", lty = 3, pch = 3, col = scol[3])
lines(1:13, resm[4, ], type = "b", lty = 4, pch = 4, col = scol[4])
abline(h = 40)
abline(h = c(50, 60, 70), lty = 3)
legend("bottomright", legend = c("elite", "middle", "worker", "NA"), lty = 1:4, pch = 1:4, col = scol)
par(oldpar)
source("R/plotED.glm.R") # From below (not eval)
```

```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm",eval = FALSE}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Women 40-64", ylim = c(0, 1))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Men 40-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswtab,cap="Table"}
vb40$social_class <- vb40$hisclass
vb40$marital_status <- vb40$civst
vb40$urban_or_rural <- vb40$urban
vb40$age_group <- vb40$age
fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb40, family = poisson)
dr <- drop1(fitt, test = "Chisq")
dr
##knitr::kable(dr)
##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm",eval = FALSE}
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
##source("R/plotED.R")
source("R/getPHglm2.R")
resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
            nper, labb)

### 2018-01-30 slides3.Rnw
first commit

@@

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
## Load remote data:
con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##load("data/vb0.rda")
load(con)
close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)

@@

\includegraphics[height=7cm]{myfigs/skum17.pdf}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}
  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}
\end{itemize}

@@

\begin{frame}{The sampling frame}

<<lexis2>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.pl, "persons"))
text(1985.2, 50, paste(n.d, "deaths"))
@
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus2,fig.width=6,fig.height = 5.9>>=

@@

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus3,fig.width=6,fig.height = 5.9>>=

##load("data/vb.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))

vbd <- vb[vb$event == 1 & vb$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

@@

<<anovatable,results='asis',echo=FALSE>>=
xtable(dr)
@

@@

{\scriptsize
<<extab,echo=FALSE,results='asis'>>=
xtable(summary(fitp)$coefficients[1:10, 1:2])
@
}

@@

\begin{frame}{Cancer mortality, ages 40-64 and 65-89}

>>>>

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
@

<<getgoing>>=
library(knitr)
library(eha)
library(xtable)
library(skum)
## Load remote data:
con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##load("data/vb0.rda")
load(con)
close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)

@@

@@

%\includegraphics[height=7cm]{myfigs/skum17.pdf}
\includegraphics[height=7cm]{myfigs/skum.pdf}
\end{column}

@@

\begin{itemize}
\item \emp{Swedish church book} records (1851--1950).
  \begin{itemize}
\item digitized by the \emp{Demographic Data Base} at Umeå University,
\begin{itemize}
  % \item \emp{\sc Popum} and
   \item \emp{\sc Poplink}
\end{itemize}
\item  \url{http://www.cedar.umu.se/}
\end{itemize}
\item The \emp{Linnaeus} data base (1960--2013)
  \begin{itemize}
\item Census data 1960, 1965, \ldots, 1990
\item Population registers ({\sc Lisa}, 1986--2013)
  \begin{itemize}
    \item \emp{Income}
    \item \emp{Education}
  \end{itemize}
 \item Death information from \emp{National Board of Health and Welfare}
   \begin{itemize}
\item Death dates from 1 January 1961 to 31 December 2013.
  \item \emp{Causes of death} ({\sc Ulorsak}, ``main cause of death'')
   \end{itemize}
\end{itemize}sampling fra
\end{itemize}

@@

\begin{frame}[fragile]{The sampling frame}

<<lexis2,echo = FALSE>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.pl, "persons"))
text(1985.2, 50, paste(n.d, "deaths"))
@

@@

\begin{frame}[fragile]{The Linnaeus data with deaths}

\begin{columns}
\begin{column}{0.6\textwidth}

<<picklexus2,fig.width=6,fig.height = 5.9,echo=FALSE>>=

@@

\begin{columns}
\begin{column}{0.6\textwidth}
<<picklexus3,fig.width=6,fig.height = 5.9,echo=FALSE>>=
##load("data/vb.rda")
##tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
##tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
tot.m <- length(unique(vb$id[vb$from != "popum" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$from != "popum" & vb$sex == "female"]))

vbd <- vb[vb$event == 1 & vb$from != "popum", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")

@@

<<anovatable,echo=FALSE>>=
knitr::kable(dr)
@

@@

{\scriptsize
<<extab,echo=FALSE>>=
knitr::kable(summary(fitp)$coefficients[1:10, 1:2])
@
}

@@

\end{document}
\begin{frame}{Cancer mortality, ages 40-64 and 65-89}
